Vector quantisation and partitioning of COVID-19 temporal dynamics in the United States

### 2020-08-02 paper/paper.tex
First round past proofing.

\title{Time series classification of COVID-19 dynamics in the United States}
\author{Chris von Csefalvay\thanks{Starschema Inc., Arlington, VA. Correspondence: \texttt{csefalvayk@starschema.net}.}}

@@

\begin{abstract}
    The statistical dynamics of a pathogen within a population depend on a range of factors: population density, the effectiveness and investment into social distancing, public policy measures and non-pharmaceutical interventions (NPIs) are only some examples of factors that influence the number of cases over time. This paper outlines a time series cluster analysis of confirmed COVID-19 cases in the United States, using a soft-DTW (Dynamic Time Warping) k-means clustering and a k-shape based clustering algorithm to identify internally consistent clusters of case counts over time. The identification of disjoint time series clusters can use past patterns to discern the future of infectious dynamics in an area, and through identifying the most likely cluster-wise trajectory, inform public health decision-making.
\end{abstract}

@@

The emergence of SARS-CoV-2, and its associated viral syndrome COVID-19, has raised important questions about the ways we analyse and identify dynamic temporal processes. In particular, by identifying similarities in principal time-dependent indicators of epidemic dynamics, such as prevalence (the number of confirmed cases over time), we can gain insight into similarities that are likely to emerge across various regions. Through this, time series clustering has the potential to play a significant role in understanding the dynamic processes that drive an outbreak.

Clustering is the wider set of algorithms within unsupervised learning that identify similar patterns among data in arbitrarily high-dimensional spaces, effectively taking a set $\mathcal{P}$ of $N$ vectors in an $n$-dimensional space, and assigning to each of these a label from the set $\mathcal{L}$, so that the assignment of each element of $\mathcal{P}$ to the groups defined by the labels comprising $\mathcal{L}$ minimise some objective function (typically referred to as the distance metric of the clustering). Cluster algorithms are widely used today and their practical applications are ubiquitous, ranging from identifying clinical phenotypes in clinical medicine\cite{ahmad2014clinical,haldar2008cluster,lochner2005cluster,weatherall2009distinct,ye2014different} through fraud detection\cite{behera2015credit,liu2013healthcare,peng2006application,sabau2012survey,subudhi2017use} to image segmentation.\cite{chuang2006fuzzy,coleman1979image,jin2018accelerating,lafata2018data,pappas1989adaptive,wu1993optimal}

Time series clustering presents a particular complication of this problem insofar as the subject of clustering is not a vector representing a single value, but rather a time series. These time series are typically not in synchrony, but rather exhibit a range of delays, lags and leads, and may depend on extrinsic hidden variables. We may formulate the essential task of time series clustering as follows. Let $X$ comprise $n$ time series $x_{1 \ldots n}$, and let $k$ denote the cardinality of the set $\mathcal{L}$ -- in other words, the number of partitions we wish to split the data into, with $k \leq n$. Then, the mapping $f: X \rightarrow l \ | \ l \in \mathcal{L}$ is a clustering if it assigns to any element $x_i \in X \ | \ i \leq n$ one (and only one) cluster $l_i \in \mathcal{L}$, so as to minimise an objective function (typically referred to in this context as a distance metric) $J$ within the cluster.

This paper examines the use of two time series clustering algorithms -- soft-DTW k-means clustering and k-shape clustering -- to identify different patterns in COVID-19 prevalence in the continental United States, and comparing the results of the classifiers for inter-classifier consistency. By isolating the barycenters of the time-shifted clusters, we can identify consistent patterns in prevalence dynamics across multiple states, quantifying the overall effect of pre-existing characteristics, population dynamics and non-pharmaceutical interventions (NPIs) between states.

@@

Given two time series $x_t: t_x \in \mathbb{Z}$ and $y_t: t_y \in \mathbb{Z}$, there exists a cost matrix $\Delta(\mathbf{x}, \mathbf{y})$ for the distance function $J$, from which we can derive the cost matrix

@@

For the two above-mentioned series, we describe the set of matrices of all possible alignments as $\mathcal{A}_{t_x, t_y}$, which is a strict subset of $\{0, 1\}^{t_x \times t_y}$. Then, DTW can be defined as the function that for any pair $(\mathbf{x}, \mathbf{y})$ identifies $A \in \mathcal{A}_{t_x, t_y}$ so as to minimise the inner product of A with the cost matrix $\Delta(\mathbf{x}, \mathbf{y})$.

@@

Thus, DTW can be conceived of as a search task, in which $\mathcal{A}_{t_x, t_y}$ is the search space within which we search for $A$ so as to minimise $\langle A, \Delta(\mathbf{x}, \mathbf{y}) \rangle$.

@@

\noindent into a single metric.\cite{janati2020spatio} Given the generalisation of the minimum metric with a smoothing factor $\gamma \geq 0$,

@@

\begin{equation}
	p = \mathop{arg min}_{r \in M} \sum_{j = 1}^n \tau(x_j, r)^q
\end{equation}

Based on dynamic time warping distances between temporal signals, we can construct a clustering that divides the COVID-19 prevalence time series for the 48 states of the contiguous United States into a number of clusters so as to minimise distances using k-nearest neighbour clustering. Soft-DTW clustering was performed using \texttt{tslearn} 0.4.1\cite{JMLR:v21:20-091} using Python 3.7, with a $\gamma$ parameter of $0.1$.

@@

k-shape clustering is a novel, robust clustering algorithm for time series that relies on iteratively refining clusters, with cross-correlation as the underlying distance metric.\cite{paparrizos2015k} Specifically, k-shape relies on a normalised version of cross-correlation, referred to in this context as Shape Base Distance (SBD): time series are Z-normalised (i.e. $\mu = 0$ and $\sigma = 1$), and the resulting cross-correlation sequence is divided by the geometric mean of the individual time series' autocorrelations. In this sense, k-shape can be understood as a k-means clustering that uses a cross-correlation based metric $SBD(\mathbf{x}, \mathbf{y})$. Let $mathbf{x}_s$ be the series $\mathbf{x}$ shifted, with zero-padding, by s, and the same be true for $mathbf{y}_s$ respectively, \emph{mutatis mutandis}. For two time series of equal length $\mathbf{x}$ and $\mathbf{y}$, we recursively define shift-wise cross-correlation for shifts in the range $s \in [-m, m]$ as

@@

Now, we can define the distance metric $SBD(\mathbf{x}, \mathbf{y})$ by

\begin{equation}
	SBD(\mathbf{x}, \mathbf{y}) = 1 - \max_w \Bigg( \frac{\hat{\rho} (\mathbf{x}, \mathbf{y})}{\sqrt{\psi_0(\mathbf{x}, \mathbf{x}) \cdot \psi_0(\mathbf{y}, \mathbf{y})}} \Bigg)
\end{equation}

@@

\noindent and as Paparrizzos and Gravano showed, Fast Fourier Transforms allow this to be calculated efficiently in $\mathcal{O}(|\mathbf{x}| log(|\mathbf{x}|))$ time rather than $\mathcal{O}(|\mathbf{x}|^2)$ time.

Similarly to the clustering effected in Subsection~\ref{sub:soft_dtw_k_means_clustering}, k-shape clustering was performed using the \texttt{tslearn} package's \texttt{clustering.KShape} classifier, with a \texttt{n\_init} setting at 16 iterations for centroid seeds, using the result with the lowest inertia, random initialization and a convergence tolerance of $10^{-6}$.

@@

The data set described in Section~\ref{sub:source_data} has been separated into three clusters. As Figure~\ref{fig:side-by-side} shows, there are three distinctly characterisable patterns based on the barycenters:

@@

@@

The geographical distribution of the permutation of k-means and k-shape classifications merits mention. As Figure

@@

@@

The geographical distribution of these is worth noting. As Figure~\ref{fig:choropleth} shows, most of the area of the continental United States currently follows the late peaking regime, and these states are currently poised to experience further growth in case counts. Only a few states (green shades) have followed an early outbreak by a significant reduction in cases and no further resurgence, which may be considered evidence of successful mitigation/suppression efforts on their part. Finally, a number of states (blue shades) have experienced early outbreaks and are exhibiting a bimodal pattern, whereby an initial surge in April to late May 2020 has been followed not by successful suppression but a reduction followed by yet another uptick in the number of reported cases of COVID-19.

As this paper has shown, time series clustering that allows for finding commonalities between time series that are by necessity out of synchrony can be helpful in illuminating geographical and regional patterns of disease dynamics. In particular, by using two different methods -- a soft-DTW based, time-shifted k-means classifier and the correlation-based k-shape classifier --, the significant consensus between such classifications has been demonstrated where the number of confirmed COVID-19 cases in the continental United States is concerned.

>>>>

\title{Vector quantisation and partitioning of COVID-19 temporal dynamics in the United States}
\author{Chris von Csefalvay\thanks{Starschema Inc., Arlington, VA. Correspondence: \texttt{csefalvayk@starschema.net}.}}

@@

\begin{abstract}
    The statistical dynamics of a pathogen within a population depend on a range of factors: population density, the effectiveness and investment into social distancing, public policy measures and non-pharmaceutical interventions (NPIs) are only some examples of factors that influence the number of cases over time by state. This paper outlines an analysis of time series vector quantisation and paritioning of COVID-19 cases in the United States, using a soft-DTW (Dynamic Time Warping) k-means clustering and a k-shape based clustering algorithm to identify internally consistent clusters of case counts over time. The identification of characteristic types of time-dependent variations can lead to the identification of patterns within sets of time series. This, in turn, can help discern the future of infectious dynamics in an area and, through identifying the most likely cluster-wise trajectory by calculating the cluster barycenter, inform public health decision-making.
\end{abstract}

@@

The emergence of SARS-CoV-2, and its associated viral syndrome COVID-19, has raised important questions about the ways we analyse and identify dynamic temporal processes. In particular, by identifying similarities in principal time-dependent indicators of epidemic dynamics, such as cumulative prevalence (the running total of confirmed cases over time), we can gain insight into similarities that are likely to emerge across various regions. Such similarities may be reflective of various hidden processes, be they related to the pathogen, to the response thereto or to various predisposing factors. By way of this, time series clustering has the potential to play a significant role in understanding the spatio-temporal factors governing the dynamic processes that drive an outbreak.

Vector quantisation and partitioning, or clustering, is the wider set of algorithms within unsupervised statistical learning that identify similar patterns among data in arbitrarily high-dimensional vector spaces, effectively taking a set $\mathcal{P}$ of $N$ vectors in an $n$-dimensional vector space and assigning to each of these a label from the label set $\mathcal{L}$, so that the assignment of each element of $\mathcal{P}$ to the groups defined by the labels comprising $\mathcal{L}$ minimise some objective function (typically referred to as the distance metric of the clustering). Cluster algorithms are widely used today and their practical applications are manifold, ranging from identifying clinical phenotypes in  medicine and population health\cite{ahmad2014clinical,haldar2008cluster,lochner2005cluster,weatherall2009distinct,ye2014different} through fraud detection\cite{behera2015credit,liu2013healthcare,peng2006application,sabau2012survey,subudhi2017use} to image segmentation.\cite{chuang2006fuzzy,coleman1979image,jin2018accelerating,lafata2018data,pappas1989adaptive,wu1993optimal}

Time series clustering presents a particular complication of this problem insofar as the subject of clustering is not a vector representing a single value, but rather a time series. These time series are typically not in synchrony, but rather exhibit a range of delays, lags and leads, and may depend on extrinsic and/or hidden variables. We may formulate the essential task of time series clustering as follows. Let $\mathcal{T}$ be a set of $n$ time series $x_{1 \ldots t_{max}}$. Further, let $k$ denote the cardinality of the label set $\mathcal{L}$ -- in other words, the number of partitions we wish to split the data into, with $k \leq n$. Then, the mapping $f: \mathcal{T} \rightarrow l \ | \ l \in \mathcal{L}$ is a clustering of the set of time series if it assigns to any element $x_i \in X \ | \ i \leq n$ one (and only one) label $l_i \in \mathcal{L}$, so as to minimise an objective function (typically referred to in this context as a distance metric) $J$ within each cluster $C_j = {x_p \in \mathcal{T} | l_p = j, j \in \mathcal{L}}$ defined by its label.

This paper examines the use of two time series clustering algorithms -- soft-DTW k-means clustering and k-shape clustering -- to identify different patterns in COVID-19 prevalence in the continental United States, and comparing the results of the classifiers for inter-classifier consistency. By isolating the barycenters of the time-shifted clusters, we can identify consistent patterns in prevalence dynamics across multiple states. This in turn can be used to quantify the overall effect of pre-existing characteristics, population dynamics and non-pharmaceutical interventions (NPIs) between states.

@@

Given two time series $\mathbf{x}: t_x \in \mathbb{Z}$ and $\mathbf{y}: t_y \in \mathbb{Z}$, there exists a cost matrix $\Delta(\mathbf{x}, \mathbf{y})$ for the distance function $J$, from which we can derive the cost matrix

@@

For the two above-mentioned series, we may describe the set of matrices of all possible alignments as $\mathcal{A}_{t_x, t_y}$, which is a strict subset of $\{0, 1\}^{t_x \times t_y}$. Then, DTW can be defined as the function that for any pair of time series $(\mathbf{x}, \mathbf{y})$ identifies an alignment $A \in \mathcal{A}_{t_x, t_y}$ so as to minimise the inner product of A with the cost matrix $\Delta(\mathbf{x}, \mathbf{y})$ as

@@

Thus, DTW can be conceived of as a search task, in which $\mathcal{A}_{t_x, t_y}$ is the search space within which we search for an alignment $A$ given $\mathbf{x}$ and $\mathbf{y}$ so as to minimise the inner product $\langle A, \Delta(\mathbf{x}, \mathbf{y}) \rangle$.

@@

\noindent into a single metric.\cite{janati2020spatio} Given the generalisation of the minimum metric with a smoothing factor $\gamma \geq 0$ as

@@

\begin{equation}
	p = \mathop{arg min}_{r \in M} \sum_{j = 1}^n \tau(x_j, r)^q \ | \ q \geq 1
\end{equation}

Thus, based on the soft-DTW metric, laid out above, we can extract from the time series of COVID-19 cumulative incidence a clustering that iteratively minimises the within-cluster sum of squares (k-means clustering). For the purposes of this paper, soft-DTW clustering was performed using \texttt{tslearn} 0.4.1\cite{JMLR:v21:20-091} using Python 3.7, with a $\gamma$ parameter of $0.1$.

@@

k-shape clustering is a novel, robust clustering algorithm for time series that relies on iteratively refining clusters, with cross-correlation as the underlying distance metric.\cite{paparrizos2015k} Specifically, k-shape relies on a normalised version of cross-correlation, referred to in this context as Shape Base Distance (SBD): time series are Z-normalised (i.e. $\mu = 0$ and $\sigma = 1$), and the resulting cross-correlation sequence is divided by the geometric mean of the individual time series' autocorrelations. In this sense, k-shape can be understood as a k-means clustering that uses a cross-correlation based metric $SBD(\mathbf{x}, \mathbf{y})$. Let $\mathbf{x}_s$ be the series $\mathbf{x}$ shifted, with zero-padding, by s, and the same be true for $\mathbf{y}_s$ respectively, \emph{mutatis mutandis}. For two time series of equal length $\mathbf{x}$ and $\mathbf{y}$, we recursively define shift-wise cross-correlation for shifts in the range $s \in [-m, m]$ as

@@

\noindent Now, we can define the distance metric $SBD(\mathbf{x}, \mathbf{y})$ by

\begin{equation}
	SBD(\mathbf{x}, \mathbf{y}) = 1 - \max_w \Bigg( \frac{\hat{\rho_w} (\mathbf{x}, \mathbf{y})}{\sqrt{\psi_0(\mathbf{x}, \mathbf{x}) \cdot \psi_0(\mathbf{y}, \mathbf{y})}} \Bigg)
\end{equation}

@@

\noindent and as Paparrizzos and Gravano showed,\cite{paparrizos2015k} Fast Fourier Transforms allow this to be calculated efficiently in $\mathcal{O}(|\mathbf{x}| log(|\mathbf{x}|))$ time rather than $\mathcal{O}(|\mathbf{x}|^2)$ time.

Similarly to the cluster analysis carried out in Subsection~\ref{sub:soft_dtw_k_means_clustering}, k-shape clustering was performed using the \texttt{tslearn} package's \texttt{clustering.KShape} implementation, with an \texttt{n\_init} setting at 16 iterations for centroid seeds, using the result with the lowest inertia, random initialization and a convergence tolerance of $10^{-6}$.

@@

After fitting the soft-DTW k-means and k-shape clustering models on the data set described in Section~\ref{sub:source_data} with a label set cardinality (i.e. number of clusters) of 3, indicators for goodness of fit were obtained using \texttt{sklearn.metrics}. These show that the clustering is relatively sound. The silhouette scores (soft-DTW k-means: 0.249, k-shape: 0.276) indicate while there is some chance of an overlap, the clustering is a relatively good fit.\cite{rousseeuw1987silhouettes} This is confirmed by a strong Variance Ratio Criterion (Calinski-Harabasz score) of 18.809 and 20.155 for soft-DTW k-means and k-shape, respectively.\cite{calinski1974dendrite} As Figure~\ref{fig:side-by-side} clearly indicates, there are three distinctly characterisable patterns based on the barycenters:

@@

Figure~\ref{fig:side-by-side} highlights in red the barycenters or Fr\^{e}chet means of the time series, expanding the notion of a centroid as a central tendency to the metric space of the time series. While the barycenters are different between the clustering algorithms (largely due to small differences in clustering, thus leading to different compositions for the barycenter calculation), they identify consistently the underlying pattern characteristic of the cluster. Notably, the barycenters calculated by the k-shape classification exhibit much stronger short-term (weekly) periodicity in all three clusters. At the same time, the second cluster's abnormal peak in June is much less reflected in the barycenter based on the k-shape clustering than it is on the k-means cluster, and the k-shape cluster presents a barycenter with a much flatter 'peak' in mid-April than the k-means barycenter.

@@

The distribution of time series (i.e. states) over the permutations of soft-DTW k-means and k-shape cluster assignments (see Figure~\ref{fig:k_combinations}) shows that the majority of states fall into matching soft-DTW k-means and k-shape categories, with only 3 states falling outside. Over half (56\%) of states fall into the soft-DTW k-means cluster 3 and k-shape cluster 3, while 8 states (17\%) fall into the soft-DTW k-means cluster 2 and k-shape cluster 2, and 9 states (21\%) fall into the soft-DTW k-means cluster 1 and k-shape cluster 1. This distribution is displayed in the inter-classifier agreement matrix in Figure~\ref{fig:inter_classifier}.

@@

The strong cross-cluster agreement, covering 96\% of all samples, indicates that despite their methodological differences, both the soft-DTW k-means clustering algorithm and the k-shape algorithm yield largely identical results when it comes to assigning states' time series to clusters. The strong concurrence and favourable ARI indicate that the cluster assignments are unlikely to be artefactual results of the underlying algorithms but rather reflect truly significantly distinct groupings of states by their case count time series.

@@

The geographical distribution of these is worth noting. As Figure~\ref{fig:choropleth} shows, at the time of writing, most of the area of the continental United States follows the late peaking regime, and the calculated barycenters indicate these states are currently poised to experience further growth in case counts. Only a few states (green shades) have followed an early outbreak with a significant reduction in cases and no further resurgence, as may be considered evidence of successful mitigation/suppression efforts on their part. Finally, a number of states (blue shades) have experienced early outbreaks and are exhibiting a bimodal pattern, whereby an initial surge in April to late May 2020 has been followed not by successful suppression but a reduction followed by yet another rise in the number of reported cases of COVID-19.

As this paper has shown, time series clustering allows for finding commonalities between time series that are by necessity out of synchrony. In doing so, it can be helpful in illuminating geographical and regional patterns of disease dynamics. In particular, by using two different methods -- a soft-DTW based, time-shifted k-means classifier and the correlation-based k-shape classifier --, a significant consensus between such classifications has been demonstrated where the number of confirmed COVID-19 cases in the continental United States is concerned. This lends credence to the hypothesis that epidemic dynamics of COVID-19 follow three distinct temporal patterns. These are in all likelihood conditioned by a combination of spatio-temporal factors (position along the epidemic's 'wavefront'), mitigation measures such as NPIs, their reltive effectiveness, as well as pre-existing factors of resilience and vulnerability.
